---
title: "The Lost Lab Notebook — Day 1"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered

---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)
library(janitor)
library(lubridate)
library(rsconnect)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(tutorial.storage = NULL)

# ── Raw data the scientist left behind ────────────────────────────────────────

day1_data <- tibble::tibble(
  experiment_id = c("exp_001", "exp_002", NA, "exp_003", "exp_004", "exp_005"),
  timestamp     = c(
    "2025-07-01 09:00:00",
    "2025-07-01 09:05:00",
    "2025-07-01 09:10:00",
    "202507-01 9-15-00",   # <‑ deliberately malformed
    "2025-07-01 09:20:00",
    "2025-07-01 09:25:00"
  ),
  comment       = c(
    "##$$",
    "TFVNRU5f",
    "1265435O27",
    "HFRSAVW7Dt",
    "####$%",
    "jiiflpa_"
  ),
  value         = c(12.5, 13.7, 14.1, 11.9, 15.2, 12.7)
)
```

---

## Introduction

this is going to be an introduction

## Day 1

###  Welcome, Investigator
You’ve unearthed a forgotten research drive from a long‑abandoned experiment. 
Inside you find a corrupted CSV: messy timestamps, cryptic notes, and—rumour has it—
a *hidden* fragment of a larger message.  
Your first mission is to clean this chaos and extract the fragment.
To do this you'll need to:

1. Clean up the mess,

1. Extract only meaningful entries,

1. Decode the comment field.



### Take a Peek
Have a look at the raw file to get your bearings.

```{r peek, exercise = FALSE}
day1_data
```

---


### Your Mission

To uncover the clue you will need to clean the data and extract the valid fragments by:

1. Keeping only rows where `experiment_id` is not missing

1. Remove rows with invalid timestamps

1. Creating a new column `fragment` that contains the comment only if it's alphanumeric (A–Z, a–z, 0–9)

1. Filtering to keep only rows where `fragment` is not missing

```{r mission, exercise = TRUE, exercise.lines = 10}
# YOUR TURN
```


```{r mission-hint-1}
# Use dplyr::filter() to remove rows where experiment_id is NA
```

```{r mission-hint-2}
# Use parse_date_time() from lubridate to convert timestamp into a datetime column
# Then filter out rows where the result is NA
```

```{r mission-hint-3}
# Use str_detect() to check if comment matches the pattern "^[A-Za-z0-9]+$"
```

```{r mission-hint-4}
# Use filter() again to remove rows with NA in the `fragment` column
```


```{r mission-solution, exercise.reveal_solution = FALSE}
final_fragments <- day1_data |> 
  filter(!is.na(experiment_id)) |> 
  mutate(
    time = parse_date_time(timestamp, orders = c("ymd HMS", "Ymd HMS")),
    fragment = ifelse(str_detect(comment, "^[A-Za-z0-9]+$"), comment, NA_character_)
  )  |> 
  filter(!is.na(time), !is.na(fragment))

final_fragments
```

---

### Submit Your Answer
What *is* the surviving fragment?

```{r question, echo = FALSE}
question_text(
  "Enter the surviving fragment:",
  answer("TFVNRU5f", correct = TRUE),
  answer("HFRSAVW7Dt", message = "Close, but not quite—double‑check your filters!"),
  allow_retry = TRUE
)
```

Congratulations! You’ve wrestled Day 1’s chaos into order and pried loose the first piece of the Lab’s secret!

